---
title: "Bootstrap"
author: "Dancun Juma"
date: "`r Sys.Date()`"
output: pdf_document
---

# Task 1: Done

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 2: Load the necessary packages
```{r include=TRUE}
library(tidyverse)
library(tidymodels)
library(GGally)
```

# Task 3: Creating the Data

```{r}
# Set a random seed value so we can obtain the same "random" results
set.seed(2025)

n <- 20

# Create a data frame/tibble named sim_dat
sim_dat <- tibble(
  # Explain what next line is doing
  x1 = runif(20, -5, 5),
  # Explain what next line is doing
  x2 = runif(20, 0, 100),
  # Explain what next line is doing
  x3 = rbinom(20, 1, 0.5)
  )

b0 <- 2
b1 <- 0.25
b2 <- -0.5
b3 <- 1
sigma <- 1.5

errors <- rnorm(20, 0, sigma)

sim_dat <- sim_dat %>%
  mutate(
    y = b0 + b1*x1 + b2*x2 + b3*x3 + errors,
    x3 = case_when(
      x3 == 0 ~ "No",
      TRUE ~ "Yes"
      )
    )
```


```{r}
ggpairs(sim_dat)
```


# Task 4: Traditional MLR Model

```{r}
mlr_fit <- linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm") %>%
  fit(y ~ x1 + x2 + x3, data = sim_dat)

tidy(mlr_fit, conf.int = TRUE)
```

# Task 5: Bootstrapping

```{r}
# Set a random seed value so we can obtain the same "random" results
set.seed(631)

# Generate the 2000 bootstrap samples
boot_samps <- sim_dat %>%
  bootstraps(times = 2000)

boot_samps
```

```{r}
# Create a function that fits a fixed MLR model to one split dataset
fit_mlr_boots <- function(split) {
  lm(y ~ x1 + x2 + x3, data = analysis(split))
}

# Fit the model to each split and store the information
boot_models <- boot_samps %>%
  mutate(
    model = map(splits, fit_mlr_boots),
    coef_info = map(model, tidy)
  )

boots_coefs <- boot_models %>%
  unnest(coef_info)

boots_coefs
```

## Bootstrap Confidence Intervals

```{r}
boot_int <- int_pctl(boot_models, statistics = coef_info, alpha = 0.05)
boot_int
```

## Visualization
```{r}
ggplot(boots_coefs, aes(x = estimate)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ term, scales = "free") +
  geom_vline(data = boot_int, aes(xintercept = .lower), col = "blue") +
  geom_vline(data = boot_int, aes(xintercept = .upper), col = "blue")
```




```{r warning=FALSE}
ggplot(boots_coefs, aes(x = estimate, fill = term)) +
  geom_histogram(bins = 30, alpha = 0.6, color = "black") +
  facet_wrap(~ term, scales = "free") +
  geom_vline(data = boot_int, aes(xintercept = .lower), col = "blue", linetype = "dashed") +
  geom_vline(data = boot_int, aes(xintercept = .upper), col = "blue", linetype = "dashed") +
  geom_vline(xintercept = c(1.5626060, 0.4336503, -0.4905565, 1.0117182), col = "orange", linetype = "solid", size = 1) +
  geom_vline(xintercept = c(2, 0.25, -0.5, 1), col = "blue", linetype = "solid", size = 1) +
  theme_minimal() +
  labs(title = "Bootstrap Estimates vs Population and Traditional Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Count")
```

# Answer to Question 5

The bootstrap estimates align closely with the population-level model coefficients. The traditional confidence intervals (orange lines) provide a reference, while the bootstrap intervals (blue dashed lines) capture the variability from resampling. 

## Accuracy Assessment:
- The bootstrap confidence intervals contain the true population values for most estimates, indicating reasonable accuracy.
- The traditional method’s confidence intervals are slightly narrower than the bootstrap intervals, reflecting the differences in estimation methods.
- The variability of bootstrap estimates is evident, but their central tendency aligns well with the expected population coefficients.

This suggests that bootstrapping provides a robust alternative to traditional inference methods, especially when assumptions about normality or small sample sizes need to be considered.

## Challenge Enhancements
- Added vertical orange lines for traditional confidence intervals.
- Added vertical solid blue lines for population slope values.
- Applied `theme_minimal()` for a cleaner plot appearance.
- Added histogram colors and adjusted transparency for better visibility.
- Included clear axis labels and a title to improve interpretability.




